### 2022-05-02 paper/localpsf.tex
minor edits

Our method forms a hierarchical matrix (H-matrix) \cite{Hackbusch99,} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high-rank (see Figure \ref{fig:hmatrix_intro} and Appendix \ref{app:h_matrix}). H-matrix methods permit us to perform matrix-vector products cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving the H-matrix that cannot easily be done using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. Typically, the work and memory required to perform these operations for an $N \times N$ H-matrix scales as $O(r^2 N \log(N)^a)$, where $r$ is the largest rank of the off-diagonal blocks, and $a \in \{0,1,2,3\}$. The exact cost depends on the type of H-matrix used and operation performed; for more details see \cite{GrasedyckHackbusch03}. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

@@

Computing $\mathbf{\Pi}=\Pi(\mathbf{A}_H^{\text{sym}})$ directly is difficult, but we may approximate $\Pi$ by the following high order rational function:
\begin{equation*}
	\Pi(\lambda) \approx \Pi_k(\lambda) := \frac{1}{1 + T(\lambda)^{2^k} }, \qquad \text{where} \qquad T(\lambda) := \frac{2 \lambda - (b+a)}{b-a},
\end{equation*}
and $k$ is a positive integer; the larger $k$ is, the better the approximation $\Pi \approx \Pi_k$. This is illustrated in Figure \ref{fig:spd_rat}. Replacing scalar operations with matrix operations, we have
\begin{equation}
\label{eq:rational343}
\mathbf{\Pi}_k := \Pi_k\left(\mathbf{A}_H^{\text{sym}}\right) = \left(\mathbf{I} + \mathbf{T}^{2^k}\right)^{-1},
\end{equation}
where
\begin{equation*}
\mathbf{T} := \left(2 \mathbf{A}_H^{\text{sym}} - (b+a) \mathbf{I}\right)/(b-a).
\end{equation*}
Using $\mathbf{\Pi}_k$ instead of $\mathbf{\Pi}$ in \eqref{eq:I_Pi_abs} allows us to efficiently approximate $|\mathbf{A}_H^{\text{sym}}|$ using H-matrix methods. 
First, we compute $\lambda_\text{min}$ using the Lanczos method, and set $a=\gamma \lambda_\text{min}$ and $b=0$, where $\gamma\ge 1$ is a parameter that must be chosen. We use $\gamma=1.5$, though the method is relatively insensitive to $\gamma$ and works well for any $\gamma \in [1,2]$. 
Next, we compute $\mathbf{T}$ using fast H-matrix scalar multiplication and addition. We form the $\mathbf{T}^{2^k}$ via repeated squaring, $\mathbf{T}^2 = \mathbf{T} \mathbf{T}$, $\mathbf{T}^4 = \mathbf{T}^2 \mathbf{T}^2$, $\mathbf{T}^8 = \mathbf{T}^4 \mathbf{T}^4$ and so on, where each of these matrix squarings is performed using fast H-matrix multiplication. We recommend using $k=1$ or $2$ (we use $k=1$ in our numerical results). Larger $k$ is unnecessary and may make the method unstable. The approximate spectral projector $\mathbf{\Pi}_k$ is formed from $\mathbf{T}^{2^k}$ per \eqref{eq:rational343}, using fast H-matrix operations to perform the required matrix-matrix addition and matrix inversion. Finally, a positive semi-definite approximation $\mathbf{A}_H^{\text{sym}+}$ of $\mathbf{A}_H^{\text{sym}}$ is formed as
\begin{equation*}
\mathbf{A}_H^{\text{sym}+} := \left(\mathbf{I} - 2\mathbf{\Pi}_k)\right)\mathbf{A}_H^{\text{sym}},
\end{equation*}
where fast H-matrix operations are used to perform the necessary matrix-matrix subtraction and multiplication. Note that for any $k \ge 1$, $\Pi_k(\lambda)> 1/2$ for $\lambda \in (a,b)$, so $\mathbf{A}_H^{\text{sym}+}$ is positive semi definite\footnote{Some indefiniteness may be introduced by truncation error in the H-matrix operations, but the negative eigenvalues caused by this truncation error are typically extremely small.}. At the end, $\mathbf{A}_H^{\text{sym}+}$ is symmetrized once again, to correct for any small non-symmetries that are introduced by truncation error in the H-matrix operations. Once constructed, we use the symmetric positive semidefinite matrix $\mathbf{A}_H^{\text{sym}+}$ in place of $\mathbf{A}_H$ in further linear algebra operations.

>>>>

Our method forms a hierarchical matrix (H-matrix) \cite{Hackbusch99,} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high-rank (see Figure \ref{fig:hmatrix_intro} and Appendix \ref{app:h_matrix}). H-matrix methods permit us to perform matrix-vector products cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving the H-matrix that cannot easily be done using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. Typically, the work and memory required to perform these operations for an $N \times N$ H-matrix scales as $O(N \log(N)^a)$, where $a \in \{0,1,2,3\}$. The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks; for more details see \cite{GrasedyckHackbusch03}. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

@@

Computing $\mathbf{\Pi}=\Pi(\mathbf{A}_H^{\text{sym}})$ directly is difficult, but we may approximate $\Pi$ by a high order rational function $\Pi_k \approx \Pi(\lambda)$ which is defined as follows:
\begin{equation*}
	 \Pi_k(\lambda) := \frac{1}{1 + T(\lambda)^{2^k} }, \qquad \text{where} \qquad T(\lambda) := \frac{2 \lambda - (b+a)}{b-a}.
\end{equation*}
Here $k$ is a positive integer; the larger $k$ is, the better the approximation $\Pi \approx \Pi_k$. This is illustrated in Figure \ref{fig:spd_rat}. Replacing scalar operations with matrix operations, we have an approximation $\mathbf{\Pi}_k \approx \mathbf{\Pi}$ given by
\begin{equation}
\label{eq:rational343}
\mathbf{\Pi}_k := \Pi_k\left(\mathbf{A}_H^{\text{sym}}\right) = \left(\mathbf{I} + \mathbf{T}^{2^k}\right)^{-1},
\end{equation}
where
\begin{equation*}
\mathbf{T} := \left(2 \mathbf{A}_H^{\text{sym}} - (b+a) \mathbf{I}\right)/(b-a).
\end{equation*}
Using $\mathbf{\Pi}_k$ instead of $\mathbf{\Pi}$ in \eqref{eq:I_Pi_abs} allows us to efficiently approximate $|\mathbf{A}_H^{\text{sym}}|$ using H-matrix methods. 
First, we compute $\lambda_\text{min}$ using the Lanczos method, and set $a=\gamma \lambda_\text{min}$ and $b=0$, where $\gamma\ge 1$ is a parameter that must be chosen. We use $\gamma=1.5$, though the method is relatively insensitive to $\gamma$ and works well for any $\gamma \in [1,2]$. 
Next, we compute $\mathbf{T}$ using fast H-matrix scalar multiplication and addition. We form the $\mathbf{T}^{2^k}$ via repeated squaring, $\mathbf{T}^2 = \mathbf{T} \mathbf{T}$, $\mathbf{T}^4 = \mathbf{T}^2 \mathbf{T}^2$, $\mathbf{T}^8 = \mathbf{T}^4 \mathbf{T}^4$ and so on, where each of these matrix squarings is performed using fast H-matrix multiplication. We recommend using $k=1$ or $2$ (we use $k=1$ in our numerical results). Larger $k$ is unnecessary and may make the method unstable. The approximate spectral projector $\mathbf{\Pi}_k$ is formed from $\mathbf{T}^{2^k}$ per \eqref{eq:rational343}, using fast H-matrix operations to perform the required matrix-matrix addition and matrix inversion. Finally, a positive semi-definite approximation $\mathbf{A}_H^{\text{sym}+}$ of $\mathbf{A}_H^{\text{sym}}$ is formed as
\begin{equation*}
\mathbf{A}_H^{\text{sym}+} := \left(\mathbf{I} - 2\mathbf{\Pi}_k)\right)\mathbf{A}_H^{\text{sym}},
\end{equation*}
where fast H-matrix operations are used to perform the necessary matrix-matrix subtraction and multiplication. Note that for any $k \ge 1$, $\Pi_k(\lambda)> 1/2$ for $\lambda \in (a,b)$, so $\mathbf{A}_H^{\text{sym}+}$ is positive semi definite\footnote{Some indefiniteness may be introduced by truncation error in the H-matrix operations, but the negative eigenvalues caused by this truncation error are typically extremely small.}. At the end, $\mathbf{A}_H^{\text{sym}+}$ is symmetrized once again, to correct for any small non-symmetries that are introduced by truncation error in the H-matrix operations. Once constructed, we use the symmetric positive semidefinite matrix $\mathbf{A}_H^{\text{sym}+}$ in place of $\mathbf{A}_H$ in further linear algebra operations.
